---
title: "RNA-Seq Analysis"
author: "Claire Chapman"
date: "11/17/2021"
output: github_document
---

## Read in Data
DESeq2 expects a dataframe of count data and a second dataframe with information about the samples (metadata)
```{r}
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("airway_metadata.csv")
```

Take a look
```{r}
head(counts)
```
```{r}
head(metadata)
```
There are `r nrow(counts)` genes in this dataset

Column names of counts data MUST equal the IDs in metadata. Check this:
```{r}
metadata$id == colnames(counts)
```


## Exploratory Diff Gene Analysis - Compare control to treated
For demonstration purposes only, never do actual diff analysis this way

Extract row from metadata with controls
```{r}
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
control.mean <- rowSums( control.counts )/4 
head(control.mean)
```


OR use dplyr (I prefer)
```{r}
library(dplyr)
control <- metadata %>% filter(dex=="control")
control.counts <- counts %>% select(control$id) 
control.mean <- rowSums(control.counts)/4
head(control.mean)
```
```{r}
treated <- metadata %>% filter(dex=="treated")
treated.counts <- counts %>% select(treated$id) 
treated.mean <- rowSums(treated.counts)/4
head(treated.mean)
```
Combine mean count data
```{r}
meancounts <- data.frame(control.mean, treated.mean)
```

## Compare the control and treated
Quick scatterplot to check work
```{r}
library(ggplot2)
meancounts %>% 
  ggplot(aes(control.mean, treated.mean)) +
  geom_point()+
  labs(x = "control", y = "treated")
```
Not truly seeing all the 60,000 data
```{r}
meancounts %>% 
  ggplot(aes(control.mean, treated.mean)) +
  scale_y_continuous(trans = "log2")+
  scale_x_continuous(trans = "log2")+
  geom_point()+
  labs(x = "control", y = "treated")
```
We often use log transformations to make life easier...

If there is no difference, log change is 0
```{r}
log2(20/20)
```
If there is twice as much, expression goes up, log change is 1
```{r}
log2(40/20)
```
If there is half as much, expression goes down, log change is -1
```{r}
log2(10/20)
```
Transform our data
Adding a column called log2 fold change representing log2(treated/control)
```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
```
```{r}
head(meancounts)
```

You get some weird values
"NaN" when you try to divide by zero
"-inf"
Need to exclude the zeros
```{r}
head(meancounts[,1:2] == 0)
```
the `which()` function tells us the indices of TRUE entries in a logical vector. It is not useful in default mode, doesn't break between columns. Need to argue with it.
```{r}
inds <- which(meancounts[,1:2] == 0, arr.ind = TRUE)
head(inds)
```
Some genes have 0 value in both columns, so select unique values
```{r}
to.rm <- unique(inds[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

We now have `r nrow(mycounts)` genes remaining.

## Up-regulation and Down-regulation
A common threshold for log2fc is up-regulated if > 2 and down-regulated if < -2.
```{r}
up_reg <- mycounts$log2fc > 2
down_reg <- mycounts$log2fc < -2
```
```{r}
sum(up_reg)
```

There are `r sum(up_reg)` genes up-regulated, or `r 100*(sum(up_reg)/nrow(mycounts))` %

There are `r sum(down_reg)` genes down-regulated or `r 100*(sum(down_reg)/nrow(mycounts))` %


## DESeq2 Analysis
```{r}
library(DESeq2)
```
```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
```
```{r}
dds
```
```{r}
dds <- DESeq(dds)
```
```{r}
res <- results(dds)
res
```

## Volcano plot
Common way to visualize the results
Add lines at our -2, 2 fc thresholds and at the significant p value threshold.
We are interested in the top left and top right quadrants.
```{r}
plot(res$log2FoldChange, -log(res$padj))
abline(v = c(-2,2), col = "gray")
abline(h = -log(0.05), col = "gray")
```

## Adding annotation data
Use Bioconductor's main annotation packages. Must first install these packages in the console.

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```
```{r}
columns(org.Hs.eg.db)
```
We want to make a new value "SYMBOL" that includes the common gene name that will be widely recognized and used.

```{r}
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res), # Our genenames
                     keytype="ENSEMBL",        # The format of our genenames
                     column="SYMBOL",          # The new format we want to add
                     multiVals="first")
head(res)
```
Now let's save it for next time!
```{r}
write.csv(res, file = "DESeq_results.csv")
```

