---
title: "Class 08"
author: "Claire Chapman"
date: "10/24/2021"
output: pdf_document
---

# Clustering methods
## kmeans()

### First lets make up some data to cluster

```{r}
tmp <- c(rnorm(30, 3), rnorm(30, -3))
tmp
```
```{r}
data <- cbind(x = tmp, y = rev(tmp))
data
```
Using just our eyes, we can assume that we should get two clusters
```{r}
plot(data)
```
### Run kmeans()
The argument **k** tells how many clusters you are asking for. The argument **nstarts** tells how many iterations 
```{r}
km <- kmeans(data, centers = 2, nstart = 20)
km
```

> Q. How many points are in each cluster?

```{r}
km$size
```
**Thirty points in each cluster**

> Q. What "component" of your result object details cluster assignment/membership?

```{r}
km$cluster
```
> Q. What "component" of your result object details cluster center?

```{r}
km$centers
```
> Q. Plot x colored by the kmeans cluster assignment and add cluster centers as blue points

```{r}
plot(data, col = km$cluster)
points(km$centers, col = "blue", pch = 15)
```

## Hierarchical Clustering, **hclust()**
A little bit more work required than in kmeans: can't just use x, need a distance matrix using function **dist()**

```{r}
hc <- hclust(dist(data))
hc
```
This printout wasn't very helpful, use a plot. Hclust() has its own useful plot method
```{r}
plot(hc)
abline(h = 7, col = "red")
```
To find out membership vector, we need to "cut" the tree. Use **cutree()** function and tell it the height to cut at
```{r}
cutree(hc, h = 7)
```
Can also use **cutree()** and state the number of k clusters we want
```{r}
grps <- cutree(hc, k = 2)
grps
```
```{r}
plot(data, col = grps)
```

# Principal Component Analysis (PCA)
A useful analysis method when you have lots of dimensions in your data

## PCA of UK food data
### Read in data
```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
```

> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
```
Why aren't there only 4 columns (one for each country)?
```{r}
head(x)
```
Row names were being assigned to the first column, incorrectly

### Fix data
```{r}
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
```
The better way to fix this is upon import
```{r}
x <- read.csv(url, row.names = 1)
dim(x)
```

> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

The second method is much better because it makes the correct change, permanently, to the dataset whereas the first method will keep taking away the first column.

### Plotting the data
```{r}
barplot(as.matrix(x), beside = T, col = rainbow(nrow(x)))
```
> Q3: Changing what optional argument in the above barplot() function results in the following plot?

The argument "beside" determines if values are stacked on top of each other or beside each other. Default is beside = F and values are stacked

> Q5: Generating all pairwise plots may help somewhat. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
mycols <- rainbow(nrow(x))
pairs(x, col=mycols, pch=16)
```
This plot compares all pairs of countries. Points on the diagonal show that tha data is the same for the pair of countries. Points that are not on the diagonal signify differences between the two countries being compared.

> Q6. What is the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

The main difference between N. Ireland and the other countries is the green data point that was consumed at a higher quantity in N.Ireland than others. The results are still just qualitative though...

## PCA is here to help
Base R function **prcomp()** to find PCA
First need to tranpose the data, this function wants countries in rows and foods in columns
```{r}
pca <- prcomp(t(x))
summary(pca)
```
```{r}
plot(pca)
```

We want the **score plot** (aka PCA plot), the plot of PC1 vs. PC2 that we often see
```{r}
attributes(pca)
```
We want pca$x component for this plot
```{r}
plot(pca$x[,1:2])
text(pca$x[,1:2], labels = colnames(x))
```

### PCA "loadings"
Loadings tell us how much each variable contributes to each PC
```{r}
par(mar = c(10, 3, 0.35, 0))
barplot(pca$rotation[,1], las = 2)
```
## One more PCA for the day
```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```

> Q10 How many genes and samples are in the data set?

```{r}
nrow(rna.data)
```
100 genes

```{r}
ncol(rna.data)
colnames(rna.data)
```
10 samples

### PCA
The argument scale normalizes the ranges in observations
```{r}
pca.rna <- prcomp(t(rna.data), scale = T)
summary(pca.rna)
```
```{r}
plot(pca.rna)
```
```{r}
plot(pca.rna$x[,1:2])
text(pca.rna$x[,1:2], labels = colnames(rna.data))
```

### Quick scree plot
```{r}
plot(pca, main = "Quick scree plot")
```

### our own scree plots 
Variance captured per PC
```{r}
pca.var <- pca.rna$sdev^2
```
Percent variance is often more informative to look at
```{r}
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
pca.var.per
```
```{r}
barplot(pca.var.per, main="Scree Plot", 
        names.arg = paste0("PC", 1:10),
        xlab="Principal Component", ylab="Percent Variation")
```
Adding colors
```{r}
colvec <- colnames(rna.data)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"

plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
     xlab=paste0("PC1 (", pca.var.per[1], "%)"),
     ylab=paste0("PC2 (", pca.var.per[2], "%)"))

text(pca$x[,1], pca$x[,2], labels = colnames(rna.data), pos=c(rep(4,5), rep(2,5)))
```

