---
title: "Class 18 HW"
author: "Claire Chapman"
date: "12/2/2021"
output: pdf_document
---

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
url <- "https://bioboot.github.io/bggn213_F21/class-material/rs8067378_ENSG00000172057.6.txt"
data <- read.table(url)
head(data)
```

```{r}
library(dplyr)
library(ggplot2)
```
```{r}
data %>% 
  group_by(geno) %>% 
  summarise(sample_size = n(), median = median(exp))
```


> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
data %>% 
  group_by(geno) %>% 
  ggplot(aes(geno, exp, fill = geno)) +
   geom_jitter(alpha = 0.5)+
  geom_boxplot()+
  labs(title = "Expression levels by genotype", x = "Genotype", y = "Expression")
```

Since there is a difference in expression between the A/A and G/G genotype, we can hypothesize that the SNP does affect the experssion of ORMDL3.