---
title: "R Functions"
author: "Claire Chapman"
date: "10/15/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> Q1. Write a function grade() to determine an overall grade from a vector of student homework
assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an
NA value) this can be used as a score to be potentially dropped. Your final function should be
adquately explained with code comments and be able to work on an example class gradebook
such as this one in CSV format: “https://tinyurl.com/gradeinput” 

Simple example vectors
```{r}
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)

sum(is.na(student3))
```

### Mean **without** dropping the lowest score
```{r}
mean(student1)
```
### Mean **with** dropping the lowest score
**which.min()** returns the position of the minimum value
```{r}
which.min(student1)
```
This will drop the lowest score from our vector
```{r}
student1[-which.min(student1)]
```
Take the mean with the lowest score dropped from the vector
```{r}
mean(student1[-which.min(student1)])
```

### Dealing with NAs
Simply removing NAs wouldn't be fair in the final grades. We want the NAs to be treated as 0s
Use **is.na()** to find and replace NAs with 0
```{r}
is.na(student2)
```
```{r}
student2[is.na(student2)] <- 0
student2
```
### Combine into one function
```{r}
x <- student1
x[is.na(x)] <- 0
mean(x[-which.min(x)])
```
```{r}
x <- student2
x[is.na(x)] <- 0
mean(x[-which.min(x)])
```
```{r}
x <- student3
x[is.na(x)] <- 0
mean(x[-which.min(x)])
```
### Time to make the function!
Go to Code > Extract function on working snippet
```{r}
grade <- function(x) {
  x[is.na(x)] <- 0
  mean(x[-which.min(x)])
}
```
```{r}
grade(student1)
```
```{r}
grade(student2)
```
```{r}
grade(student3)
```

### Add code comments
For better understanding of **why** in the future
```{r}
grade <- function(x) {
  # make missing homework scores equal to zero
  x[is.na(x)] <- 0
  # drop the lowest homework score
  mean(x[-which.min(x)])
}
```
Code > Insert Roxygen Skeleton to make your own "Help" page for your new function
```{r}
#' Calculate average score for a vector of homework scores while dropping the lowest score. Missing values are treated as 0 scores
#'
#' @param x Numeric vector of homework scores
#'
#' @return Average score
#' @export
#'
#' @examples
#'  student <- c(100, NA, 90, 80)
#'  grade(student)
grade <- function(x) {
  # make missing homework scores equal to zero
  x[is.na(x)] <- 0
  # drop the lowest homework score
  mean(x[-which.min(x)])
}
```

### Apply function to entire gradebook
Read in the spreadsheet, replace the quotes that were in the wrong font
```{r}
url <- "https://tinyurl.com/gradeinput"
gradebook <- read.csv(url)
```
Change first column of students to row names
```{r}
gradebook <- read.csv(url, row.names = 1)
```
Use **apply()** to apply the **grade()** function to the whole gradebook
"1" denotes going by row
```{r}
apply(gradebook, 1, grade)
```

> Q2. Using your grade() function and the supplied gradebook, Who is the top scoring student
overall in the gradebook?

Use **sort** to pick out the top score from the results
```{r}
results <- apply(gradebook, 1, grade)
sort(results, decreasing = TRUE)
```
Another way to find the top score
```{r}
which.max(results)
```

**Student 18 has the top average score**


> Q3. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained
the lowest scores overall?

Find the median (less affected by outliers) of each column of the gradebook to see which assignment had the lowest performance
```{r}
toughest_hw <- apply(gradebook, 2, median, na.rm=TRUE)
which.min(toughest_hw)
```
**Homework 2 was the toughest on students when using the median**

Always a good idea to check with a plot of your data
```{r}
boxplot(gradebook)
```


> Q4. Optional Extension: From your analysis of the gradebook, which homework was most
predictive of overall score (i.e. **highest correlation with average grade score**)?

Use cor() to find the best correlation between the results and the gradebook
```{r}
gradebook[is.na(gradebook)] <- 0
correlations <- apply(gradebook, 2, cor, x = results)
```
```{r}
which.max(correlations)
```
**Homework 5 scores were the most correlated with overall performance by students**

