---
title: 'Class 17: Vaccination Mini Project'
author: "Claire Chapman"
date: "11/28/2021"
output: html_document
---

## Getting Started
Load your data
```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```
> Q1. What column details the total number of people fully vaccinated?

8

> Q2. What column details the Zip code tabulation area?

2

> Q3. What is the earliest date in this dataset?

2021-01-05

> Q4. What is the latest date in this dataset?

2021-11-23

```{r}
skimr::skim(vax)
```
> Q5. How many numeric columns are in this dataset?

9

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column?

8355

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?

8355/82908 = 10.08%

> Q8. [Optional]: Why might this data be missing?

Because the CDC was unable to collect the data from the given zip code due to missing infrastructure or noncompliance

### Working with dates
```{r}
library(lubridate)
```
```{r}
today()
```
Convert to a format that lubridate will understand 
```{r}
vax$as_of_date <-  ymd(vax$as_of_date)
```

> Q9. How many days have passed since the last update of the dataset?

```{r}
today() - vax$as_of_date[nrow(vax)]
```

> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)?

```{r}
unique(vax$as_of_date)
```
47 unique dates


### Working with ZIP codes
```{r}
library(zipcodeR)
```

```{r}
geocode_zip('92037')
```
Calculate distance between two zip code centers
```{r}
zip_distance('92037', '92109')
```
Pull census data
```{r}
reverse_zipcode(c('92037', "92109"))
```

## Focus on the San Diego area
```{r}
library(dplyr)
```
```{r}
sd <- vax %>% 
  filter(county == "San Diego")
```

> Q11. How many distinct zip codes are listed for San Diego County?

```{r}
unique(sd$zip_code_tabulation_area)
```
107 zip codes in San Diego County

> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?

```{r}
sd %>% 
  filter(age12_plus_population == max(sd$age12_plus_population))
```
92154 zip code has the age 12 plus population of 76365.2

> Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2021-11-09”?

```{r}
sd %>% 
  filter(as_of_date == "2021-11-09") %>% 
  summarise(avg_fullvax = mean(percent_of_population_fully_vaccinated, na.rm = TRUE))
```


> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2021-11-09”?

```{r}
library(ggplot2)
```


```{r}
nov_9 <- sd %>% 
  filter(as_of_date == "2021-11-09")
hist(nov_9$percent_of_population_fully_vaccinated)
```

## Focus on UCSD/La Jolla
```{r}
ucsd <- filter(sd, zip_code_tabulation_area == "92037")
```

> Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area

```{r}
ucsd %>% 
  ggplot(aes(x = as_of_date, y = percent_of_population_fully_vaccinated)) +
  geom_point()+
  geom_line(group = 1)+
  ylim(c(0,1)) +
  labs(title = "Vaccination Rate for La Jolla CA", y = "Percent Vaccinated", x = "Date")
```
> Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
sd %>% 
  filter(as_of_date == "2021-11-16") %>% 
  filter(age5_plus_population >= 36144) %>% 
  summarise(mean_vaxrate = mean(percent_of_population_fully_vaccinated))
```
```{r}
ucsd %>% 
  ggplot(aes(x = as_of_date, y = percent_of_population_fully_vaccinated)) +
  geom_point()+
  geom_line(group = 1)+
  ylim(c(0,1)) +
  labs(title = "Vaccination Rate for La Jolla CA", y = "Percent Vaccinated", x = "Date")+
  geom_hline(yintercept = 0.6744255)
```
> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”?

```{r}
vax %>% 
  filter(as_of_date == "2021-11-16") %>% 
  filter(age5_plus_population >= 36144) %>% 
  summarise(min = min(percent_of_population_fully_vaccinated), median = median(percent_of_population_fully_vaccinated), mean = mean(percent_of_population_fully_vaccinated), sec_qtr = median/2, third_qtr = 3/2*(median),  max = max(percent_of_population_fully_vaccinated))
```
> Q18. Using ggplot generate a histogram of this data.

```{r}
vax %>% 
  filter(as_of_date == "2021-11-16") %>% 
  filter(age5_plus_population >= 36144) %>% 
  ggplot(aes(x = percent_of_population_fully_vaccinated)) +
  geom_histogram()
```

> Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```
Lower in ZIP 92040

```{r}
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```
Greater in ZIP 92109

> Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144

```{r}
vax.36 <- vax %>% 
  filter(age5_plus_population > 36144) 
vax.36 %>% 
  ggplot(aes(x = as_of_date,  y = percent_of_population_fully_vaccinated, group = zip_code_tabulation_area)) +
  geom_line(alpha = 0.2)+
  ylim(c(0,1))+
  labs(x = "Date", y = "Vaccination Rate", title = "Vaccination Rates Across California", subtitle = "In ZIP codes with populations greater than 36k")+
  geom_hline(yintercept = 0.6744255)
```
> Q21. How do you feel about traveling for Thanksgiving and meeting for in-person class next Week?

I know the decision has already been made but thanks for thinking about our health after the traveling!
