---
title: "Class 16 RNA Seq Mini Project"
author: "Claire Chapman"
date: "11/20/2021"
output: pdf_document
---

## Differential Expression Analysis
Download the data
```{r}
library(DESeq2)
```

```{r}
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"
```

Take a look at metadata
```{r}
colData = read.csv(metaFile, row.names = 1)
head(colData)
```
Take a look at countData
```{r}
countData = read.csv(countFile, row.names = 1)
head(countData)
```
We do not want the first column of the count data called "length". All columns must be the same as the rows of our meta data.
```{r}
countData <- as.matrix(countData[,-1])
head(countData)
```
Let's get rid of the zero data so it doesn't mess up future calculations
```{r}
countData = countData[-which(rowSums(countData)==0),]
head(countData)
```
### Running DESeq2
Set up the DESeqDataSet object required for the function
```{r}
dds = DESeqDataSetFromMatrix(countData = countData, colData = colData, design =~condition)
dds = DESeq(dds)
```
```{r}
dds
```
```{r}
res = results(dds)
```

Use summary function to get a feel of how many genes are up or down regulated
```{r}
summary(res)
```
### Volcano Plot
```{r}
plot(res$log2FoldChange, -log(res$padj))
```
Add code to polish the graph
```{r}
mycols <- rep("gray", nrow(res) )

# genes with absolute fold change above 2 will be red
mycols[abs(res$log2FoldChange) > 2] <- "red"

# genes with adjusted p-value less than 0.01 and absolute fold change more than 2 will be blue
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2)
mycols[inds] <- "blue"

plot(res$log2FoldChange, -log(res$padj), col = mycols, xlab = "Log2(FoldChange)", ylab = "-Log(P-value)" )

# add some cut off lines
abline(v = c(-2,2), col = "gray", lty = 2)
abline(h = -log(0.1), col = "gray", lty = 2)
```
Adding some gene annotation
```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```
```{r}
columns(org.Hs.eg.db)
```
```{r}
res$symbol = mapIds(org.Hs.eg.db, keys = row.names(res), keytype = "ENSEMBL", column = "SYMBOL", multiVals = "first")

res$entrez = mapIds(org.Hs.eg.db, keys = row.names(res), keytype = "ENSEMBL", column = "ENTREZID", multiVals = "first")

res$name = mapIds(org.Hs.eg.db, keys = row.names(res), keytype = "ENSEMBL", column = "GENENAME", multiVals = "first")

head(res, 10)
```
Reorder by p-value and save them to our current directory
```{r}
res = res[order(res$pvalue),]
write.csv(res, file = "deseq_results.csv")
```

## Pathway Analysis
Time to use the **gage** pathway for pathway analysis. First we find a list of enriched pathways, then we use **pathview** to draw pathway diagrams

Install one time only in console
```{r}
library(pathview)
library(gage)
library(gageData)

data(kegg.sets.hs)
data(sigmet.idx.hs)
```

Narrow down to signaling and metabolic pathways only
```{r}
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]
head(kegg.sets.hs, 3)
```
**gage()** requires a named vector of fold changes (from DESeq2 analysis) with names of values as Entrez gene IDs (obtained from mapIDs())
```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```
Use Gage
```{r}
keggres = gage(foldchanges, gsets = kegg.sets.hs)
```
```{r}
attributes(keggres)
head(keggres$less)
```
```{r}
pathview(gene.data = foldchanges, pathway.id = "hsa04110")
```
Another way to present the data...
```{r}
pathview(gene.data = foldchanges, pathway.id = "hsa04110", kegg.native = FALSE)
```
Let's try to pull out the top 5 upregulated pathways to use for future pathview plotting

Focus on top 5
```{r}
keggrespathways <- rownames(keggres$greater)[1:5]
```
Extract IDs of each string
```{r}
keggresids = substr(keggrespathways, start = 1, stop = 8)
keggresids
```
Now use these IDs in pathview to show paths for the top 5
```{r}
pathview(gene.data = foldchanges, pathway.id = keggresids, species = "hsa")
```
Try to do this but with the 5 most downregulated genes
Focus on top 5
```{r}
keggrespathways.down <- rownames(keggres$less)[1:5]
```

Extract the IDs (8 characters) of each string
```{r}
keggresids.down <- substr(keggrespathways.down, start = 1, stop = 8)
keggresids.down
```
View the data
```{r}
pathview(gene.data = foldchanges, pathway.id = keggresids.down, species = "hsa", low = "blue", mid = "green", high = "yellow")
```


